---
title: "Statistical analysis: Diversity"
output:
  github_document:
    toc: true
editor_options:
  chunk_output_type: console
---

## Read in packages and data
```{r message=FALSE, warning=FALSE, error=TRUE}
library(tidyverse)
library(dplyr)
library(caret)
library(ggplot2)
library(phyloseq)
library(vegan)
library(lme4)
library(lmerTest)
library(car)
library(ggpubr)
library(ggprism)

packageVersion("tidyverse")
packageVersion("dplyr")
packageVersion("caret")
packageVersion("ggplot2")
packageVersion("phyloseq")
packageVersion("vegan")
packageVersion("lme4")
packageVersion("lmerTest")
packageVersion("car")
packageVersion("ggpubr")
packageVersion("ggprism")

path <- "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/Mikrobiom korrelation/"

```

Read in data:
```{r}
ps_count <- readRDS(paste0(path, "data/IBD_project/ps_total_count.rds"))
ps_relab <- readRDS(paste0(path, "data/IBD_project/ps_total_relab.rds"))

alpha_div_ASV <- read.csv(paste0(path, "data/IBD_project/AlphaDiv_ASV.txt"), sep="\t")
```

## Analyze alpha diversity

### Analysis of technical features
Technical features: Study-ID, instrument, single/paired, read depth, V-region
Test whether the number of detected ASVs and genera are depending on these features.
```{r}
#Remove the samples also removed from the ps object due to less than 2000 genus reads pr. sample:
alpha_div_ASV <- alpha_div_ASV %>% filter(Run %in% sample_data(ps_count)$Run)
alpha_div_ASV <- alpha_div_ASV %>% filter(!(BioProject %in% c("PRJNA606913", "PRJNA313074", "PRJNA436359", "PRJNA761255")))

#Check whether data should be transformed:
ggplot(alpha_div_ASV)+
  geom_histogram(aes(x=Chao1))+theme_classic()+ylab("Count")
p1 <- ggplot(alpha_div_ASV)+
  geom_histogram(aes(x=log(Chao1+1)))+theme_classic()+ylab("Count")+xlab("Chao1 - log transformed")+theme(text = element_text(family = "sans", size = 8))

p2 <- ggplot(alpha_div_ASV)+
  geom_histogram(aes(x=Shannon))+theme_classic()+ylab("Count")+theme(text = element_text(family = "sans", size = 8))

p_total <- ggarrange(p1, p2, labels = c("A)", "B)"), font.label = list(size=10, family="sans"))

ggsave(p_total, filename = paste0(path, "illustrations/IBD_project/FigureS3.tiff"), width = 174, units = "mm", dpi=300)

alpha_div_ASV$Chao1_transformed <- log(alpha_div_ASV$Chao1 +1)

#Add read depth from ps object:
meta_2 <- as.data.frame(as.matrix(sample_data(ps_count))) %>% dplyr::select(Run, N)
meta_2$N <- as.numeric(meta_2$N)
alpha_div_ASV <- merge(alpha_div_ASV, meta_2, by.x="Run", by.y="Run")

#Transform N to be on similar scale as alpha_div
ggplot(alpha_div_ASV)+
  geom_histogram(aes(x=N))

ggplot(alpha_div_ASV)+
  geom_histogram(aes(x=log(N+1)))

alpha_div_ASV$N_transformed <- log(alpha_div_ASV$N+1)

#Check how many outliers of N:
check <- alpha_div_ASV %>% filter(!is.na(Disease)&!is.na(Host_Age)&!is.na(host_sex)&!is.na(geo_loc_name_country_continent))

sum(check$N_transformed>mean(check$N_transformed)+3*sd(check$N_transformed)| check$N_transformed<mean(check$N_transformed)-3*sd(check$N_transformed))
8/2518

sum(check$N>mean(check$N)+3*sd(check$N)| check$N<mean(check$N)-3*sd(check$N))
48/2518

#Remove mean +- 3*SD - alpha diversity:
alpha_sub_Chao1 <- alpha_div_ASV %>% filter(Chao1_transformed < mean(alpha_div_ASV$Chao1_transformed)+3*sd(alpha_div_ASV$Chao1_transformed)) %>% filter(Chao1_transformed > mean(alpha_div_ASV$Chao1_transformed)-3*sd(alpha_div_ASV$Chao1_transformed)) 
#5339 samples

alpha_sub_Shannon <- alpha_div_ASV %>% filter(Shannon < mean(alpha_div_ASV$Shannon)+3*sd(alpha_div_ASV$Shannon)) %>% filter(Shannon > mean(alpha_div_ASV$Shannon)-3*sd(alpha_div_ASV$Shannon)) 
#5347 samples


# Test all technical features together in a linear model:
f_asv <- lm(Chao1_transformed ~ Instrument + LibraryLayout + X16s_region+N_transformed , data=alpha_sub_Chao1)
car::vif(f_asv)
summary(f_asv)
f_1 <- aov(f_asv)
car::Anova(f_1, type = 2)

# Repeat for Shannon:
f_asv <- lm(Shannon ~ Instrument + LibraryLayout + X16s_region + N_transformed, data=alpha_sub_Shannon)
car::vif(f_asv)
summary(f_asv)
f_1 <- aov(f_asv)
car::Anova(f_1, type = 2)
```


### Analysis of biological covariates
Biological features: Geography, gender, age, BMI, smoking - test them one at the time, because the choice will reduce the sample size a lot
Include important technical features as covariates
```{r}
alpha_sub_Chao1$Host_Age <- as.numeric(alpha_sub_Chao1$Host_Age)
alpha_sub_Shannon$Host_Age <- as.numeric(alpha_sub_Shannon$Host_Age)
alpha_sub_Chao1$host_body_mass_index <- as.numeric(alpha_sub_Chao1$host_body_mass_index)
alpha_sub_Shannon$host_body_mass_index <- as.numeric(alpha_sub_Shannon$host_body_mass_index)

#Remove extreme (and definitely wrongly notated) BMIs
alpha_sub_Chao1$host_body_mass_index[as.numeric(alpha_sub_Chao1$host_body_mass_index) < 10] <- NA
alpha_sub_Chao1$host_body_mass_index[as.numeric(alpha_sub_Chao1$host_body_mass_index) > 100] <- NA
alpha_sub_Shannon$host_body_mass_index[as.numeric(alpha_sub_Shannon$host_body_mass_index) < 10] <- NA
alpha_sub_Shannon$host_body_mass_index[as.numeric(alpha_sub_Shannon$host_body_mass_index) > 100] <- NA

#Chao1, ASV 
alpha_sub_Chao1$technical <- paste0(alpha_sub_Chao1$Instrument, "_", alpha_sub_Chao1$LibraryLayout, "_", alpha_sub_Chao1$X16s_region)

#Sex
f_asv <- lm(Chao1_transformed ~ technical + N_transformed + host_sex, data=alpha_sub_Chao1, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
f_1 <- aov(f_asv)
car::Anova(f_1, type = 2)
sum(!is.na(alpha_sub_Chao1$host_sex)) #3172 samples

#Age
f_asv <- lm(Chao1_transformed ~ technical + N_transformed + Host_Age, data=alpha_sub_Chao1, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
sum(!is.na(alpha_sub_Chao1$Host_Age)) #2615 samples

#Country
f_asv <- lm(Chao1_transformed ~ technical + N_transformed + geo_loc_name_country, data=alpha_sub_Chao1, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
f_1 <- aov(f_asv)
car::Anova(f_1, type = 2)
sum(!is.na(alpha_sub_Chao1$geo_loc_name_country)) #5223 samples

#Continent
f_asv <- lm(Chao1_transformed ~ technical + N_transformed + geo_loc_name_country_continent, data=alpha_sub_Chao1, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
f_1 <- aov(f_asv)
car::Anova(f_1, type = 2)

#BMI
f_asv <- lm(Chao1_transformed ~technical + N_transformed + log(host_body_mass_index+1), data=alpha_sub_Chao1, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
sum(!is.na(alpha_sub_Chao1$host_body_mass_index)) #1784 samples


#Smoking
f_asv <- lm(Chao1_transformed ~ technical + N_transformed + smoking_status, data=alpha_sub_Chao1, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
f_1 <- aov(f_asv)
car::Anova(f_1, type = 2)
sum(!is.na(alpha_sub_Chao1$smoking_status)) #1831 samples

#Shannon, ASV 
alpha_sub_Shannon$technical <- paste0(alpha_sub_Shannon$Instrument, "_", alpha_sub_Shannon$LibraryLayout, "_", alpha_sub_Shannon$X16s_region)
#Sex
f_asv <- lm(Shannon ~ technical + N_transformed + host_sex, data=alpha_sub_Shannon, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
f_1 <- aov(f_asv)
car::Anova(f_1, type = 2)
sum(!is.na(alpha_sub_Shannon$host_sex)) #3171 samples

#Age
f_asv <- lm(Shannon ~ technical + N_transformed+ Host_Age, data=alpha_sub_Shannon, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
sum(!is.na(alpha_sub_Shannon$Host_Age)) #2611 samples

#Country
f_asv <- lm(Shannon ~ technical + N_transformed+ geo_loc_name_country, data=alpha_sub_Shannon, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
f_1 <- aov(f_asv)
car::Anova(f_1, type = 2)
sum(!is.na(alpha_sub_Shannon$geo_loc_name_country)) #5225 samples

#Continent
f_asv <- lm(Shannon ~ technical + N_transformed+ geo_loc_name_country_continent, data=alpha_sub_Shannon, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
f_1 <- aov(f_asv)
car::Anova(f_1, type = 2)

#BMI
f_asv <- lm(Shannon ~ technical + N_transformed+ log(host_body_mass_index+1), data=alpha_sub_Shannon, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
sum(!is.na(alpha_sub_Shannon$host_body_mass_index)) #1767 samples

#Smoking
f_asv <- lm(Shannon ~ technical + N_transformed+ smoking_status, data=alpha_sub_Shannon, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
f_1 <- aov(f_asv)
car::Anova(f_1, type = 2)
sum(!is.na(alpha_sub_Shannon$smoking_status)) #1812 samples

```


### Analysis of the association between microbiome composition and disease status
Include important techincal and biological features as covariates

Main analysis:
```{r}
alpha_sub_Chao1$Disease <- factor(alpha_sub_Chao1$Disease, levels = c("HC", "CD", "UC"))
alpha_sub_Shannon$Disease <- factor(alpha_sub_Shannon$Disease, levels = c("HC", "CD", "UC"))

alpha_sub_Chao1$technical <- paste0(alpha_sub_Chao1$Instrument, "_", alpha_sub_Chao1$X16s_region, "_", alpha_sub_Chao1$LibraryLayout, "_", alpha_sub_Chao1$geo_loc_name_country_continent)
alpha_sub_Shannon$technical <- paste0(alpha_sub_Shannon$Instrument, "_", alpha_sub_Shannon$X16s_region, "_", alpha_sub_Shannon$LibraryLayout, "_", alpha_sub_Shannon$geo_loc_name_country_continent)


#Chao1 - main analysis
f_asv <- lm(Chao1_transformed ~ N_transformed + technical + Host_Age + Disease, data=alpha_sub_Chao1, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
exp(-0.4088173)-1
exp(-0.2694306)-1

#Shannon - main analysis and sensitivity (BMI and smoking was not significant)
f_asv <- lm(Shannon ~ N_transformed + technical + Host_Age + Disease, data=alpha_sub_Shannon, na.action = na.exclude)
vif(f_asv)
summary(f_asv)


#Check: is there a significant difference between UC and CD?
alpha_sub_Chao1$Disease <- factor(alpha_sub_Chao1$Disease, levels = c("UC", "HC", "CD"))
alpha_sub_Shannon$Disease <- factor(alpha_sub_Shannon$Disease, levels = c("UC", "HC", "CD"))

#Chao1 - main analysis
f_asv <- lm(Chao1_transformed ~ N_transformed + technical + Host_Age + Disease, data=alpha_sub_Chao1, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
exp(-0.1393866)-1

#Shannon, ASV
f_asv <- lm(Shannon ~ N_transformed + technical + Host_Age + Disease, data=alpha_sub_Shannon, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
```

Sensitivity analysis:
```{r}

alpha_sub_Chao1$Disease <- factor(alpha_sub_Chao1$Disease, levels = c("HC", "CD", "UC"))
alpha_sub_Shannon$Disease <- factor(alpha_sub_Shannon$Disease, levels = c("HC", "CD", "UC"))

#Chao1:
f_asv <- lm(Chao1_transformed ~ N_transformed + technical + Host_Age +host_body_mass_index+ smoking_status + Disease, data=alpha_sub_Chao1, na.action = na.exclude)
vif(f_asv)
summary(f_asv)

#Shannon:
f_asv <- lm(Shannon ~ N_transformed + technical + Host_Age+host_body_mass_index+ smoking_status + Disease, data=alpha_sub_Shannon, na.action = na.exclude)
vif(f_asv)
summary(f_asv)


#Check: is there a significant difference between UC and CD?
alpha_sub_Chao1$Disease <- factor(alpha_sub_Chao1$Disease, levels = c("UC", "HC", "CD"))
alpha_sub_Shannon$Disease <- factor(alpha_sub_Shannon$Disease, levels = c("UC", "HC", "CD"))

#Chao1
f_asv <- lm(Chao1_transformed ~ N_transformed + technical+host_body_mass_index+ smoking_status + Host_Age + Disease, data=alpha_sub_Chao1, na.action = na.exclude)
vif(f_asv)
summary(f_asv)

#Shannon, ASV
f_asv <- lm(Shannon ~ N_transformed + technical+host_body_mass_index+ smoking_status + Host_Age + Disease, data=alpha_sub_Shannon, na.action = na.exclude)
vif(f_asv)
summary(f_asv)
```


## Ploting of results
Alpha diversity:
```{r}
#Main: 
project_list <- read.csv(paste0(path, "Metadata/Project_list.txt"), sep="\t")
alpha_sub_Chao1 <- merge(alpha_sub_Chao1, project_list, by.x="BioProject", by.y="Project")
alpha_sub_Shannon <- merge(alpha_sub_Shannon, project_list, by.x="BioProject", by.y="Project")

alpha_sub_Chao1$Disease1 <- ifelse(alpha_sub_Chao1$Disease=="HC", "HI", ifelse(alpha_sub_Chao1$Disease=="CD", "CD", "UC"))
alpha_sub_Shannon$Disease1 <- ifelse(alpha_sub_Shannon$Disease=="HC", "HI", ifelse(alpha_sub_Shannon$Disease=="CD", "CD", "UC"))

alpha_sub_Chao1$Disease1 <- factor(alpha_sub_Chao1$Disease1, levels = c("HI", "UC", "CD"))
alpha_sub_Shannon$Disease1 <- factor(alpha_sub_Shannon$Disease1, levels = c("HI", "UC", "CD"))

Chao1_sub <- alpha_sub_Chao1 %>% filter(!is.na(Disease)&!is.na(Host_Age)&!is.na(host_sex)&!is.na(geo_loc_name_country_continent))

Shannon_sub <- alpha_sub_Shannon %>% filter(!is.na(Disease)&!is.na(Host_Age)&!is.na(host_sex)&!is.na(geo_loc_name_country_continent))

colorBlindGrey8   <- c("#009E73", "#56B4E9", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")


p1 <- ggplot(data=Chao1_sub)+
  geom_boxplot(aes(x=Author, y=Chao1_transformed, color=Disease1), position = position_dodge(preserve = "single"))+theme_classic() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), text = element_text(family = "sans", size = 8))+xlab("")+ylab("Chao1 (log transformed)")+labs(color="Disease")+scale_color_manual(values=colorBlindGrey8)


ggplot(data=alpha_sub_Shannon)+
  geom_boxplot(aes(x=Author, y=Shannon, color=Disease1))+theme_classic() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+xlab("")

# Not divided by projects:
Chao1_sub$Disease1 <- factor(Chao1_sub$Disease1, levels = c("HI", "UC", "CD"))
Shannon_sub$Disease1 <- factor(Shannon_sub$Disease1, levels = c("HI", "UC", "CD"))

p_val_ref <- tibble::tribble(
  ~group1, ~group2, ~p.adj, ~y.position, 
  "HI", "CD", "< 2e-16", 7.4, 
  "HI", "UC", "< 2e-16", 7.0,
  "CD", "UC", "1.3e-4", 7.0
)

p2_1 <- ggplot(data=Chao1_sub)+
  geom_boxplot(aes(x=Disease1, y=Chao1_transformed, color=Disease1))+theme_classic()+ theme(legend.position="none")+xlab("") + ylab("Chao1 richness (log transformed)")+add_pvalue(p_val_ref, fontfamily = "sans", label.size = 2.3)+scale_color_manual(values=colorBlindGrey8)+theme(text = element_text(family = "sans", size = 8))

p_val_ref2 <- tibble::tribble(
  ~group1, ~group2, ~p.adj, ~y.position, 
  "HI", "CD", "4.4e-13", 7.6, 
  "HI", "UC", "5.0e-06", 7,
  "CD", "UC", "0.364", 7
)

p2_2 <- ggplot(data=Shannon_sub)+
  geom_boxplot(aes(x=Disease1, y=Shannon, color=Disease1))+theme_classic()+ theme(legend.position="none")+xlab("") + ylab("Shannon diversity")+add_pvalue(p_val_ref2, fontfamily = "sans", label.size = 2.3)+scale_color_manual(values=colorBlindGrey8)+theme(text = element_text(family = "sans", size = 8))

p2 <- ggarrange(p2_1, p2_2)
``` 



## Analyze beta diversity
### Analysis of technical features

The beta diversity technical part is run on the super-computer, so outputs will not be shown -> it takes a long time to run! The results can be found in a separate folder. 
```{r eval=FALSE, include=TRUE}
ps_relab <- readRDS(paste0(path, "data/IBD_project/ps_total_relab.rds"))
#ps_relab <- readRDS("./vol1/ps_total_relab.rds")

df1 <- data.frame(sample_data(ps_relab))
df1$N <- as.numeric(df1$N)

#No permutation design based on BioProject! Because within the same BioProject, they will all have the same technical variables -> cannot permute within and test technical

set.seed(1)
#Bray Curtis:
res1 <- adonis2(otu_table(ps_relab)~ Instrument + LibraryLayout + X16s_region + N, data=df1, method="bray", by="margin")
write.table(res1, file = "./vol1/BrayCurtis_technical_margin.txt")

set.seed(1)
#Jaccard:
res2 <- adonis2(otu_table(ps_relab)~ Instrument + LibraryLayout + X16s_region + N, data=df1, method="jaccard", by="margin", binary=T)
write.table(res2, file = "./vol1/Jaccard_technical_margin.txt")

set.seed(1)
#Bray Curtis:
res3 <- adonis2(otu_table(ps_relab)~ Instrument + LibraryLayout + X16s_region + N, data=df1, method="bray", by=NULL)
write.table(res3, file = "./vol1/BrayCurtis_technical_full.txt")

set.seed(1)
#Jaccard:
res4 <- adonis2(otu_table(ps_relab)~ Instrument + LibraryLayout + X16s_region + N, data=df1, method="jaccard", binary=T, by = NULL)
write.table(res4, file = "./vol1/Jaccard_technical_full.txt")
```

### Analysis of biological covariates
Technical effects are not included as covariates, as there is already corrected for them in the permutation design.
```{r cache=TRUE}
#Permutation design: Within, but not between projects
#Here, it is okay to use adonis instead of adonis2 -> we will just use the sequential estimates since order doesn't matter, when there is just one variable (will save a lot of time!). 
ps_relab <- readRDS(paste0(path, "data/IBD_project/ps_total_relab.rds"))

#Sex
ps_relab_sub <- subset_samples(ps_relab, !is.na(host_sex))
df1 <- data.frame(sample_data(ps_relab_sub))
df1$N <- as.numeric(df1$N)
perm <- how(within=Within(type="free"), plots=Plots(strata=df1$BioProject, type="none"), nperm=999)

set.seed(1)
adonis(otu_table(ps_relab_sub)~ N+host_sex, permutations = perm, data=df1, method="bray")

set.seed(1)
adonis(otu_table(ps_relab_sub)~ N+host_sex, permutations = perm, data=df1, method="jaccard", binary=T)

#Age
ps_relab_sub <- subset_samples(ps_relab, !is.na(Host_Age))
df1 <- data.frame(sample_data(ps_relab_sub))
df1$Host_Age <- as.numeric(df1$Host_Age)
df1$N <- as.numeric(df1$N)
perm <- how(within=Within(type="free"), plots=Plots(strata=df1$BioProject, type="none"), nperm=999)

set.seed(1)
adonis(otu_table(ps_relab_sub)~ N+Host_Age, permutations = perm, data=df1, method="bray")

set.seed(1)
adonis(otu_table(ps_relab_sub)~ N+Host_Age, permutations = perm, data=df1, method="jaccard", binary=T)

#Country
ps_relab_sub <- subset_samples(ps_relab, !is.na(geo_loc_name_country))
df1 <- data.frame(sample_data(ps_relab_sub))
#Try something else with country/continent -> almost does not differ within the same project!
df1$N <- as.numeric(df1$N)
set.seed(1)
adonis(otu_table(ps_relab_sub)~ Instrument + LibraryLayout + X16s_region + N + geo_loc_name_country, data=df1, method="bray")
set.seed(1)
adonis(otu_table(ps_relab_sub)~ Instrument + LibraryLayout + X16s_region + N + geo_loc_name_country, data=df1, method="jaccard", binary=T)
set.seed(1)
adonis(otu_table(ps_relab_sub)~ Instrument + LibraryLayout + X16s_region + N + geo_loc_name_country_continent, data=df1, method="bray")
set.seed(1)
adonis(otu_table(ps_relab_sub)~ Instrument + LibraryLayout + X16s_region + N + geo_loc_name_country_continent, data=df1, method="jaccard", binary=T)

#BMI
ps_relab_sub <- subset_samples(ps_relab, !is.na(host_body_mass_index))
ps_relab_sub <- subset_samples(ps_relab_sub, host_body_mass_index<100)
ps_relab_sub <- subset_samples(ps_relab_sub, host_body_mass_index>10)
df1 <- data.frame(sample_data(ps_relab_sub))
df1$host_body_mass_index <- as.numeric(df1$host_body_mass_index)
df1$N <- as.numeric(df1$N)
perm <- how(within=Within(type="free"), plots=Plots(strata=df1$BioProject, type="none"), nperm=999)

set.seed(1)
adonis(otu_table(ps_relab_sub)~ N+host_body_mass_index, permutations = perm, data=df1, method="bray")

set.seed(1)
adonis(otu_table(ps_relab_sub)~ N+host_body_mass_index, permutations = perm, data=df1, method="jaccard", binary=T)

#Smoking status
ps_relab_sub <- subset_samples(ps_relab, !is.na(smoking_status))
df1 <- data.frame(sample_data(ps_relab_sub))
df1$N <- as.numeric(df1$N)
perm <- how(within=Within(type="free"), plots=Plots(strata=df1$BioProject, type="none"), nperm=999)

set.seed(1)
adonis(otu_table(ps_relab_sub)~ N+smoking_status, permutations = perm, data=df1, method="bray")

set.seed(1)
adonis(otu_table(ps_relab_sub)~ N+smoking_status, permutations = perm, data=df1, method="jaccard", binary=T)
```

### Analysis of the association between microbiome composition and disease status

Main analysis:
Also here, we can use adonis instead of adonis2. Just make sure, that Disease is in the end, so we capture the effect of disease after the effects of covariates are adjusted for!
```{r cache=TRUE}

ps_relab_sub <- subset_samples(ps_relab, !is.na(Disease))
ps_relab_sub <- subset_samples(ps_relab_sub, !is.na(Host_Age))
ps_relab_sub <- subset_samples(ps_relab_sub, !is.na(host_sex))
ps_relab_sub <- subset_samples(ps_relab_sub, !is.na(geo_loc_name_country)) #2518 samples

df1 <- data.frame(sample_data(ps_relab_sub))
df1$Host_Age <- as.numeric(df1$Host_Age)
df1$N <- as.numeric(df1$N)

perm <- how(within=Within(type="free"), plots=Plots(strata=df1$BioProject, type="none"), nperm=999)

set.seed(1)
#Bray Curtis:
adonis(otu_table(ps_relab_sub)~ N+Host_Age + host_sex + geo_loc_name_country + Disease, data=df1, method="bray", permutations=perm)

set.seed(1)
#Jaccard:
adonis(otu_table(ps_relab_sub)~ N+ Host_Age + host_sex + geo_loc_name_country + Disease, data=df1, method="jaccard", permutations=perm, binary=T)

#Divide into HC+CD and HC+UC to test separately:
ps_UC <- subset_samples(ps_relab_sub, Disease !="CD") #1827 samples
ps_CD <- subset_samples(ps_relab_sub, Disease !="UC") #2275 samples

#Test first CD vs. HC:
df1 <- data.frame(sample_data(ps_CD))
df1$Host_Age <- as.numeric(df1$Host_Age)
df1$N <- as.numeric(df1$N)

perm <- how(within=Within(type="free"), plots=Plots(strata=df1$BioProject, type="none"), nperm=999)
set.seed(1)
#Bray Curtis:
adonis(otu_table(ps_CD)~ N+Host_Age + host_sex + geo_loc_name_country + Disease, data=df1, method="bray", permutations=perm)

set.seed(1)
#Jaccard:
adonis(otu_table(ps_CD)~  N+Host_Age + host_sex + geo_loc_name_country + Disease, data=df1, method="jaccard", permutations=perm, binary=T)

#Then UC vs. HC:
df1 <- data.frame(sample_data(ps_UC))
df1$Host_Age <- as.numeric(df1$Host_Age)
df1$N <- as.numeric(df1$N)

perm <- how(within=Within(type="free"), plots=Plots(strata=df1$BioProject, type="none"), nperm=999)
set.seed(1)
#Bray Curtis:
adonis(otu_table(ps_UC)~ N+Host_Age + host_sex + geo_loc_name_country + Disease, data=df1, method="bray", permutations=perm)

set.seed(1)
#Jaccard:
adonis(otu_table(ps_UC)~ N+ Host_Age + host_sex + geo_loc_name_country + Disease, data=df1, method="jaccard", permutations=perm, binary=T)
```

Sensitivity analysis:
```{r cache=TRUE}
ps_relab_sub <- subset_samples(ps_relab, !is.na(Disease))
ps_relab_sub <- subset_samples(ps_relab_sub, !is.na(Host_Age))
ps_relab_sub <- subset_samples(ps_relab_sub, !is.na(host_sex))
ps_relab_sub <- subset_samples(ps_relab_sub, !is.na(geo_loc_name_country)) 
ps_relab_sub <- subset_samples(ps_relab_sub, !is.na(host_body_mass_index))
ps_relab_sub <- subset_samples(ps_relab_sub, host_body_mass_index<100)
ps_relab_sub <- subset_samples(ps_relab_sub, host_body_mass_index>10)
ps_relab_sub <- subset_samples(ps_relab_sub, !is.na(smoking_status)) #1602 samples!

df1 <- data.frame(sample_data(ps_relab_sub))
df1$Host_Age <- as.numeric(df1$Host_Age)
df1$host_body_mass_index <- as.numeric(df1$host_body_mass_index)
df1$N <- as.numeric(df1$N)

perm <- how(within=Within(type="free"), plots=Plots(strata=df1$BioProject, type="none"), nperm=999)

set.seed(1)
#Bray Curtis:
adonis(otu_table(ps_relab_sub)~ N+Host_Age + host_sex + host_body_mass_index +smoking_status + geo_loc_name_country + Disease, data=df1, method="bray", permutations=perm)

set.seed(1)
#Jaccard:
adonis(otu_table(ps_relab_sub)~  N+Host_Age + host_sex + host_body_mass_index +smoking_status + geo_loc_name_country + Disease, data=df1, method="jaccard", permutations=perm, binary=T)

#Divide into HC+CD and HC+UC to test separately:
ps_UC <- subset_samples(ps_relab_sub, Disease !="CD") #1504 samples
ps_CD <- subset_samples(ps_relab_sub, Disease !="UC") #1471 samples

#Test first CD vs. HC:
df1 <- data.frame(sample_data(ps_CD))
df1$Host_Age <- as.numeric(df1$Host_Age)
df1$host_body_mass_index <- as.numeric(df1$host_body_mass_index)
df1$N <- as.numeric(df1$N)

perm <- how(within=Within(type="free"), plots=Plots(strata=df1$BioProject, type="none"), nperm=999)
set.seed(1)
#Bray Curtis:
adonis(otu_table(ps_CD)~ N+Host_Age + host_sex+ host_body_mass_index +smoking_status  + geo_loc_name_country + Disease, data=df1, method="bray", permutations=perm)

set.seed(1)
#Jaccard:
adonis(otu_table(ps_CD)~  N+Host_Age + host_sex + host_body_mass_index +smoking_status + geo_loc_name_country + Disease, data=df1, method="jaccard", permutations=perm, binary=T)

#Then UC vs. HC:
df1 <- data.frame(sample_data(ps_UC))
df1$Host_Age <- as.numeric(df1$Host_Age)
df1$host_body_mass_index <- as.numeric(df1$host_body_mass_index)
df1$N <- as.numeric(df1$N)

perm <- how(within=Within(type="free"), plots=Plots(strata=df1$BioProject, type="none"), nperm=999)
set.seed(1)
#Bray Curtis:
adonis(otu_table(ps_UC)~ N+Host_Age + host_sex + host_body_mass_index +smoking_status + geo_loc_name_country + Disease, data=df1, method="bray", permutations=perm)

set.seed(1)
#Jaccard:
adonis(otu_table(ps_UC)~  N+Host_Age + host_sex + host_body_mass_index +smoking_status + geo_loc_name_country + Disease, data=df1, method="jaccard", permutations=perm, binary=T)
```

## Plot results
```{r cache=TRUE}
ps_relab_sub <- subset_samples(ps_relab, !(is.na(Disease)))
to_hellinger <- as.data.frame(otu_table(ps_relab_sub))
to_hellinger <- vegan::decostand(to_hellinger, method = "hellinger")
to_hellinger[1:5, 1:5]
ps_relab_plot <- phyloseq(otu_table(to_hellinger, taxa_are_rows = F), sample_data(ps_relab_sub), tax_table(ps_relab_sub))

#Bray Curtis: 
beta_div_relab_bray = phyloseq::distance(ps_relab_plot, method='bray')
beta_div_relab_bray %>% as.vector %>% summary

#Jaccard:
beta_div_relab_jaccard = phyloseq::distance(ps_relab_plot, method='jaccard', binary = TRUE)
beta_div_relab_jaccard %>% as.vector %>% summary
```

Plot NMDS:
```{r cache=TRUE}
#Bray Curtis
set.seed(0)
BRAY_NMDS_relab=metaMDS(beta_div_relab_bray, k=2,trymax=30)
BRAY_NMDS_relab$stress #0.2046969

data.scores = as.data.frame(scores(BRAY_NMDS_relab))
data.scores$Disease <- sample_data(ps_relab_sub)$Disease
data.scores$Disease <- ifelse(data.scores$Disease=="HC", "HI", data.scores$Disease)
data.scores$Disease <- factor(data.scores$Disease, levels = c("HI", "UC", "CD"))

p3_1 <- ggplot(data=data.scores, aes(x=NMDS1, y=NMDS2, color=Disease))+
  geom_point(size=0.7)+theme_classic()+stat_ellipse(linetype="dashed", size=1.5)+ggtitle("Bray-Curtis dissimilarity")+theme(text = element_text(family = "sans", size = 8))+scale_color_manual(values = colorBlindGrey8)


#Jaccard:
data.scores_jac <- read.csv( paste0(path, "data/IBD_project/beta_plot_Jaccard.txt"))
JAC_NMDS_relab=metaMDS(beta_div_relab_jaccard,k=2,trymax=30)
JAC_NMDS_relab$stress #0.1963084

data.scores_jac = as.data.frame(scores(JAC_NMDS_relab))
data.scores_jac$Disease <- sample_data(ps_relab_sub)$Disease
data.scores_jac$Disease <- ifelse(data.scores_jac$Disease=="HC", "HI", data.scores_jac$Disease)
data.scores_jac$Disease <- factor(data.scores_jac$Disease, levels = c("HI", "UC", "CD"))

p3_2 <- ggplot(data=data.scores_jac, aes(x=NMDS1, y=NMDS2, color=Disease))+
  geom_point(size=0.7)+theme_classic()+stat_ellipse(linetype="dashed", size=1.5)+ggtitle("Jaccard dissimilarity")+theme(text = element_text(family = "sans", size = 8))+scale_color_manual(values = colorBlindGrey8)


write.csv(data.scores, paste0(path, "data/IBD_project/beta_plot_BrayCurtis.txt"), quote = F)
write.csv(data.scores_jac, paste0(path, "data/IBD_project/beta_plot_Jaccard.txt"), quote = F)

data.scores <- read.csv(paste0(path, "data/IBD_project/beta_plot_BrayCurtis.txt"))
data.scores_jac <- read.csv(paste0(path, "data/IBD_project/beta_plot_Jaccard.txt"))
```

```{r}
p_top <- ggarrange(p1, p2, ncol=2, labels = c("A)", "B)"), font.label = list(size=10, family="sans"), widths = c(1, 0.8))

p_bottom <- ggarrange(p3_1, p3_2, common.legend =T, legend = "right", labels="C)", font.label = list(size=10, family="sans"))

p_total <- ggarrange(p_top, p_bottom, nrow=2)

ggsave(p_total, filename = paste0(path, "illustrations/IBD_project/Figure2.tiff"), width = 174, height = 145, units = "mm", dpi=300)
```

